---
title: "Long Beach Animal Shelter - EDA"
author: "Miha Likar"
format: html
editor: visual
editor_options: 
  chunk_output_type: inline
---

```{r}
library(tidyverse)
library(ggplot2)
library(scales)
library(paletteer)
library(treemapify)
library(knitr)
library(sjPlot)
library(performance)
library(broom)
```

# Introduction

In this report, I'll present exploratory data analysis of Long Beach Animal Shelter dataset. Additionally, throughout the EDA, I will formulate specific research question(s) which will be addressed in the regression models section.

The workflow and structure of this report is as follows:

-   Brief introduction about the dataset and available variables
-   Explore distributions of various single variables
-   Explore associations between 2 and/or more variables
-   Impute certain variables based on the available data (age, reproductive status)
-   Clearly formulate RQ and Hs
-   Regression analysis (model comparison)

# Exploratory data analysis (EDA)

## Reading the dataset

Read the data from an online source - TidyTuesday GitHub page.

```{r}
longbeach_raw <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-03-04/longbeach.csv')
```

Let's look at the dataset, to see what we are working with.

```{r}
glimpse(longbeach_raw)
```

There are several variables that already stand out as variables we want to primarily focus on in our EDA. These are:

-   `animal_type`
-   `sex`
-   `dob` + `intake_date`
-   `intake_condition`
-   `intake_type`
-   `outcome_date`
-   `outcome_type`

Before we proceed, let's factorize all the character variables. If we realize we need some of them in the numeric or character format later on, we can change this afterwards.

```{r}
longbeach <- longbeach_raw %>% mutate(across(where(is.character), as.factor))
```

Ok let's look at the summary, get the basic idea of levels of categorical variables etc.

```{r}
summary(longbeach)
```

**Looking at the data, I notice a few things, that I should do, focus on:**

-   `animal_type`: lump, investigate distribution.
-   `sex`: we can use this variable to infer both sex of the animal and "reproductive state" - whether animal was neutered/spayed, or it animal's reproductive functioning is intact. Ivestigate distribution.
-   Use `dob` (date of birth) and `intake_date` to compute age at intake. This might be useful for modeling later on, but initially look at the distribution.
-   Investigate `intake_condition`. Probably categories will need lumping. Explore distribution.
-   Investigate `intake_type`. Probably categories will need lumping. Explore distribution.
-   Investigate `outcome_type`. Probably categories will need lumping. Explore distribution. Group the general types into *positive outcomes* and *negative outcomes* (will be discussed in more details later) and *adoption rate*, based on adopted vs. all other outcome types.

## Exploring distributions of variables of interest

### Animal type

Let's first check the levels of the variable `animal_type` and their count, and consider possible lumping strategies.

```{r}
longbeach %>%
  count(animal_type)
```

While it is possible that we would use lumping immediately to identify the most frequent animal types (either by fixed number of levels or proportion of the level vs. total count), we can also merge them manually based on reasoning and supporting *literature*.

It was reasonable to expect that cats and dogs will be the most prevalent categories. They are followed by birds. Category wild follows and category other is tightly behind. It seems legitimate to keep these levels as they are at this point. Rabbits and guinea pigs could be collapsed into category "Small mammals". Reptiles and amphibians could be collapsed into "Exotic", a type frequently used by shelters, pet stores etc. Livestock can definitely be merged with other, considering the group size of the livestock.

```{r}
longbeach <- longbeach %>%
  mutate(animal_type = fct_collapse(animal_type,
                                   Dogs = "dog",
                                   Cats = "cat",
                                   Birds = "bird",
                                   Wildlife = "wild",
                                   `Small mammals` = c("rabbit", "guinea pig"),
                                   Exotic = c("reptile","amphibian"),
                                   Other = c("other", "livestock")))

```

These are the categories that we will work with throughout EDA. I can anticipate that perhaps only dogs and cats will be features in the regression models due to significant discrepancy in sub-sample size of these two categories, compared to smaller ones such as birds, wildlife etc. Dataset documentation does not provide precise information about the group "Other". It is possible that various different species are included in this type.

```{r}
longbeach %>% 
  count(animal_type, sort = TRUE)
```

We can use this variable for some basic distribution/count visualisation now.

```{r}
library(jcolors)
library(paletteer)
library(ggpomological)

longbeach %>% 
  count(animal_type) %>%
  mutate(animal_type = fct_reorder(animal_type, n)) %>% 
  ggplot(aes(x = animal_type, y = n, fill = animal_type)) + 
  geom_col() + 
  theme_minimal() + 
  coord_flip() + 
  labs(x = NULL, y = "Number of animals", title = "Number of animals taken in by animal type (Long Beach Animal Shelter, 2017–2024)", caption = "Small mammals include rabbits and guneia pigs. Exotic includes reptiles and amphibians.", fill = "Animal type") + 
  scale_y_continuous(breaks = seq(0,15000,2500)) + 
  #scale_fill_brewer(palette = "Spectral", guide = guide_legend(reverse = TRUE)) +
  #scale_fill_dutchmasters(palette = "milkmaid") + 
  #scale_fill_pomological() + 
  #scale_fill_paletteer_d("PrettyCols::Autumn") +
  #scale_fill_jcolors(palette = "pal7") + 
  scale_fill_paletteer_d("ggthemes::Tableau_10") +
  theme(plot.title.position = "plot",
        plot.caption = element_text(hjust = -0.5),
        legend.position="none")
# Consider removing the legend

```

We can play around a little bit and try so called "lollipop" visualization, as it has been said to be a more modern, appealing version of bar charts.

```{r}
longbeach %>% 
  count(animal_type) %>%
  mutate(animal_type = fct_reorder(animal_type, n)) %>%
  ggplot(aes(x = n, y = animal_type)) + 
  geom_segment(aes(x = 0, xend = n, yend = animal_type, color = animal_type), linewidth = 3) +
  geom_point(aes(color = animal_type), size = 10) +
  geom_text(aes(label = n), color = "#313131", size = 2.5) + 
  labs(y = NULL, x = "Number of animals", title = "Number of animals taken in by animal type (Long Beach Animal Shelter, 2017–2024)", caption = "Small mammals include rabbits and guneia pigs. Exotic includes reptiles and amphibians.", fill = "Animal type") + 
  scale_x_continuous(breaks = seq(0,15000,2500)) + 
  theme_minimal() +
  theme(plot.title.position = "plot",
        plot.caption = element_text(hjust = -0.5),
        legend.position="none") + 
  scale_color_paletteer_d("ggthemes::Tableau_10")
```

I want to play around with different visualizations a bit and thought that while bar chart might be very simple to read and informative for both an advanced as well as lay reader, I can try out treemaps.

```{r}
library(treemapify)

longbeach %>%
  count(animal_type) %>%
  mutate(label_size = rescale(n, to = c(8, 60))) %>%  # scale size between 3 and 8
  ggplot(aes(area = n, fill = animal_type, label = paste0(animal_type, "\n", n))) +
  geom_treemap() +
  geom_treemap_text(aes(size = label_size), colour = "#474747", place = "centre", grow = FALSE) +
  scale_size_identity() +
  scale_fill_paletteer_d("ggthemes::Tableau_10") + 
  labs(title = "Number of animals taken in by animal type \n(Long Beach Animal Shelter, 2017–2024)", subtitle = "Area of each type represents proportion of the total intake.", caption = "Small mammals include rabbits and guneia pigs. Exotic includes reptiles and amphibians.") +
  theme_minimal() + 
  theme(plot.title.position = "plot",
        plot.caption = element_text(hjust = 0.05),
        legend.position="none") +
  coord_fixed(1)
```

Fine for now. We can always return to these graphs and polish the appearance. We can now apply the same logic to investigate other variables!

###Intake condition

Let's first check the levels of the variable `intake_condition` and their count, and consider possible lumping strategies.

```{r}
longbeach %>%
  count(intake_condition, sort = TRUE)
```

Huh, alright, there are 17 different conditions. It appears that we could collapse illness and injury together in "physical condition". Let's explore lumping a bit, with fixed number of categories and proportions of total sample.

```{r}
#Fixed number of categories
longbeach %>%
  mutate(intake_condition = fct_lump(intake_condition, n = 10)) %>%
  count(intake_condition, sort = TRUE)

#Number of categories based on proportions
longbeach %>%
  mutate(intake_condition = fct_lump(intake_condition, prop = .03)) %>%
  count(intake_condition, sort = TRUE)
```

**Based on category inspection I decided the following:**

-   merge illness and injury in "physical" category
-   I am also considering joining with behavioral (mild, moderate, severe) or collapsing these categories into "other" as the number of observations is really small
-   There seem to be some specific smaller categories (aged, welfare seizures etc.), which will be collapsed into "other".

We need to collapse the categories manually.

```{r}
longbeach <- longbeach %>% 
  mutate(intake_condition = fct_lump(intake_condition, 10)) %>%
  mutate(intake_condition = fct_collapse(intake_condition,
                                         Normal = "normal",
                                         `Mild physical` = c("ill mild", "injured  mild"),
                                         `Moderate physical` = c("ill moderatete", "injured  moderate"),
                                         `Severe physical` = c("ill severe", "injured  severe"),
                                         `Under age/weight` = "under age/weight",,
                                         other_level = "Other"))

longbeach %>% 
  count(intake_condition, sort = TRUE)
```

First, I will specify colors associated with each category, so they remain the same throughout visualisations.

```{r}
intake_colors <- c(
  "Normal" = "#77b7b2",
  "Mild physical" = "#ebcb2a",
  "Moderate physical" = "#f28d2b",
  "Severe physical" = "#e15859",
  "Under age/weight" = "#c8868d",
  "Other" = "#cbc8c5"
)
```

Let's now use same visualizations as before for `animal_type` to plot distribution of `intake_condition`.

```{r}
longbeach %>% 
  count(intake_condition) %>%
  mutate(intake_condition = fct_reorder(intake_condition, n)) %>% 
  ggplot(aes(x = intake_condition, y = n, fill = intake_condition)) + 
  geom_col() + 
  theme_minimal() + 
  coord_flip() + 
  labs(x = "Intake condition", y = "Number of animals", title = "Number of animals taken in, by intake condition \n(Long Beach Animal Shelter, 2017–2024)", caption = "Physical conditions include illnesses and injuries.", fill = "Intake condition") + 
  scale_y_continuous(breaks = seq(0,15000,2500)) + 
  scale_fill_manual(values = intake_colors) +
  theme(plot.title.position = "plot",
        plot.caption = element_text(hjust = -0.3),
        legend.position="none")
# Consider removing the legend

```

We can play around a little bit and try so called "lollipop" visualization, as it has been said to be a more modern, appealing version of bar charts.

```{r}
longbeach %>% 
  count(intake_condition) %>%
  mutate(intake_condition = fct_reorder(intake_condition, n)) %>%
  ggplot(aes(x = n, y = intake_condition)) + 
  geom_segment(aes(x = 0, xend = n, yend = intake_condition, color = intake_condition), linewidth = 4) +
  geom_point(aes(color = intake_condition), size = 11) +
  geom_text(aes(label = n), color = "#313131",  size = 2.5) + 
  labs(y = NULL, x = "Number of animals", title = "Number of animals taken in, by intake condition \n(Long Beach Animal Shelter, 2017–2024)", caption = "Physical conditions include illnesses and injuries.", fill = "Animal type") + 
  scale_x_continuous(breaks = seq(0,15000,5000)) + 
  theme_minimal() +
  theme(plot.title.position = "plot",
        plot.caption = element_text(hjust = -0.3),
        legend.position="none") + 
  scale_color_manual(values = intake_colors)
```

I want to play around with different visualizations a bit and thought that while bar chart might be very simple to read and informative for both an advanced as well as lay reader, I can try out treemaps.

```{r}
library(treemapify)

longbeach %>%
  count(intake_condition) %>%
  mutate(label_size = rescale(n, to = c(10, 40))) %>%  # scale size between 3 and 8
  ggplot(aes(area = n, fill = intake_condition, label = paste0(intake_condition, "\n", n))) +
  geom_treemap() +
  geom_treemap_text(aes(size = label_size), colour = "#474747", place = "centre", grow = FALSE) +
  scale_size_identity() +
  scale_fill_manual(values = intake_colors) + 
  labs(title = "Number of animals taken in, by intake condition \n(Long Beach Animal Shelter, 2017–2024)", subtitle = "Area of each intake condition represents proportion of the total intake.", caption = "Physical conditions include illnesses and injuries.") +
  theme_minimal() + 
  theme(plot.title.position = "plot",
        plot.caption = element_text(hjust = 0.05),
        legend.position="none") +
  coord_fixed(1)
```

From these plots we can see that the majority of the animals taken into the Long Beach Animal Shelter are in normal physical condition. About half less is animals that are under age or under weight. This is roughly compared to all three categories of hindered physical condition combined \~ mild, moderate, severe. *At this point I start thinking that intake condition might be an interesting predictor in the regression models. At this point we can plot `animal_type` and `intake_condition` together and explore in what conditions are different animals usually taken into the shelter.*

#### Combining animal type and intake condition

For each type of animal, what number/proportion of them arrived in each intake condition?

```{r}
longbeach %>% 
  mutate(animal_type = fct_relevel(animal_type, "Other", "Wildlife", "Exotic", "Small mammals", "Birds", "Dogs", "Cats")) %>% 
  mutate(intake_condition = fct_relevel(
  intake_condition,
  "Other",
  "Severe physical",
  "Moderate physical",
  "Mild physical",
  "Under age/weight",
  "Normal")) %>%
  ggplot(aes(x = animal_type, fill = intake_condition)) + 
  geom_bar() + 
  theme_minimal() + 
  labs(x = "Animal type", 
       y = "Number of animals",
       title = "Number of animals with different intake conditions for each animal type \n(Long Beach Animal Shelter, 2017–2024)",
       fill = "Intake condition") + 
  scale_fill_manual(values = intake_colors, guide = guide_legend(reverse = TRUE))
```

While this visualization is informative about absolute number of observations of each intake condition for different animal types. We again see that cats, followed by dogs dominate in overall number of intakes. Interesting observation is that there is quite a large proportion of cats taken in under-aged or under-weight condition. Majority of dogs on the other hand appear to be accepted in normal condition. It might also be good to visualize proportions within each group, to get an insight into more specific between-animal type differences, that might remain hidden in column-plotting with huge number of observations between the groups.

```{r}
longbeach %>% 
  filter(animal_type != "livestock") %>% 
  mutate(animal_type = fct_relevel(animal_type, "Other", "Wildlife", "Exotic", "Small mammals", "Birds", "Dogs", "Cats")) %>% 
  mutate(intake_condition = fct_relevel(
  intake_condition,
  "Severe physical",
  "Moderate physical",
  "Mild physical",
  "Other",
  "Under age/weight",
  "Normal")) %>%
  ggplot(aes(x = animal_type, fill = intake_condition)) + 
  geom_bar(position = "fill") + 
  theme_minimal() + 
  labs(x = "Animal type", 
       y = "Proportion of the specific intake condition (%)",
       title = "Proportion of specific intake condition for each animal type \n(Long Beach Animal Shelter, 2017–2024)",
       fill = "Intake condition") + 
  coord_flip() + 
  scale_y_continuous(
  labels = label_percent(), 
  breaks = seq(0, 1, 0.2)) + 
  scale_fill_manual(values = intake_colors, guide = guide_legend(reverse = TRUE))
```

OK, let's try a slightly different approach and tweak the plot by adding percentage scale on both side of the axis, order animal types based on percentage of normal intake condition and add total number of counts next to animal names.

```{r}
# Let's first pull the order of the names by proportion of normal intake condition - I tried doing it in the same pipeline, but it does not work

animal_order <- longbeach %>%
  count(animal_type, intake_condition) %>%
  group_by(animal_type) %>%
  mutate(type_n = sum(n),
         p = n / sum(n),
         animal_type_lab = paste0(animal_type, " (", type_n, ")")) %>%
  filter(intake_condition == "Normal") %>%
  arrange(p) %>%
  pull(animal_type_lab)



longbeach %>% 
  count(animal_type, intake_condition) %>%
  group_by(animal_type) %>%
  mutate(type_n = sum(n)) %>%
  ungroup() %>%
  mutate(animal_type_lab = paste0(animal_type, " (", type_n, ")"),
         animal_type_lab = factor(animal_type_lab, levels = animal_order),
         intake_condition = fct_relevel(intake_condition,
                                        "Severe physical",
                                        "Moderate physical",
                                        "Mild physical",
                                        "Other",
                                        "Under age/weight",
                                        "Normal")) %>%
  ggplot(aes(x = animal_type_lab, y = n, fill = intake_condition)) + 
  geom_col(position = "fill") + 
  theme_minimal() + 
  labs(x = "Animal type", 
       y = "Proportion of the specific intake condition (%)",
       title = "Proportion of specific intake condition for each animal type \n(Long Beach Animal Shelter, 2017–2024)", subtitle = "Numbers next to the animal types represent number of individuals from that group.",
       fill = "Intake condition") + 
  coord_flip() + 
  scale_y_continuous(
  labels = label_percent(), 
  breaks = seq(0, 1, 0.2),
  sec.axis = dup_axis(labels = label_percent(), name = NULL)) + 
  scale_fill_manual(values = intake_colors, guide = guide_legend(reverse = TRUE))
```

Considering this visualisation, there are few things we can think about:

-   It seems that exotic animals and small mammals usually arrive/are accepted in normal condition. Possibly as they might escape from home for a short while and are registered into the system upon arrival, however owners potentially pick them up soon. Thus these are not animals that would linger the streets for a long time.
-   Dogs are admitted to the shelter in normal condition much more frequently compared to cats.
-   However, cats have really high proportion of under-aged or under-weight intakes. It is possible, that stray cats give birth prematurely more often compared to dogs. On the other hand, it is possible that survival on the streets is the hardest for cats, thus leaving a lot of members of the stray cat community under-fed and under-weight.
-   Birds and wildlife have high proportion of (severe) physical conditions. Perhaps a lot of members of these groups arrive to the shelter because of the injury in the first place. It is possible that wild animals roam into urban environment or that birds have all sorts of accidents and are then injured brought to the shelter by members of the community or taken in by members of the shelter.

### Intake type

Based on the points raised in the end of the previous section, we can consider looking into `intake_type` variable.

```{r}
longbeach %>% 
  count(intake_type, sort = TRUE)
```

Majority of the animals are either stray or come from wildlife, and a noticeable proportion of animals being surrendered by the owner. Due to highly unbalanced groups, I will not explore this variable for now, will just check the count for each type of intake for different types of animals and visualize it.

```{r}
longbeach %>%
  count(animal_type, intake_type) %>%
  mutate(animal_type = fct_relevel(animal_type,
                                   "Cats", "Dogs", "Small mammals", "Exotic")) %>%
  ggplot() + 
  aes(animal_type, n, fill = intake_type) %>%
  geom_col(position = "fill") + 
  #geom_col() + 
  theme_minimal() + 
  labs(y = "Proportion of animals", x = "Animal type", fill = "Intake type", title = "Proportion of animals by different intake types for each type of animal \n(Long Beach Animal Shelter, 2017–2024)") + 
  scale_fill_paletteer_d("ggthemes::Classic_Green_Orange_12")
```

Majority of dogs and cats in particular are stray animals and a noticeable percentage of animals in those groups coming by owner's surrender. Birds, wildlife and other species are predominantly found in wild life, and presumably injured or ill, as discussed in the end of the previous section.

### Sex

Let's now proceed with the `sex` variable. During initial data inspection, I noticed that there are not only sex information, but also whether animal was neutered/spayed and is thus not sexually active/capable of reproduction anymore. As I am still seriously considering modeling adoption rate in regression models, this information might be valuable. Let's first look into the variable, mutate it and visualize.

```{r}
longbeach %>%
  count(sex)
```

OK, I will leave category Unknown as it is. Then, I will split the original `sex` variable into two, one about sex and one about reproductive state. From the meaning of the words, we can infer then when the condition is "Neutered" it refers to the male, and when the condition is "Spayed", it refers to the females. After looking into online forum discussion and consulting various AIs, I concluded that when reproductive status is listed, we can infer the sex. When it is not, we can conclude that the reproductive state is intact and biological sex is listed.

Based on this I will create binary variable `starilised`, which will be 1 if current `sex` is "Spayed" or "Neutered" and 0 if biological sex is listed. I will recode values in the original `sex` category to only include information about biological sex.

```{r}
longbeach <- longbeach %>%
  mutate(sterilised = case_when(sex %in% c("Spayed", "Neutered") ~ "Yes",
                                sex %in% c("Male", "Female") ~ "No",
                                sex == "Unknown" ~ "Unknown"),
         sex = case_when(sex %in% c("Spayed", "Female") ~ "Female",
                                sex %in% c("Male", "Neutered") ~ "Male",
                                sex == "Unknown" ~ "Unknown"))
longbeach %>%
  count(sex, sterilised) %>%
  ggplot() +
  aes(sex, n, fill = sterilised) + 
  geom_col(position = "dodge") +
  #geom_col(position = "fill") + 
  labs(x = "Sex", y = "Number of animals", fill = "Sterilised", title = "NUmber of sterilised and unstarilised animals by sex \n(Long Beach Animal Shelter, 2017–2024)") + 
  theme_minimal() + 
  scale_fill_pomological() 
```

```{r}
longbeach %>%
  count(sex, sterilised)
```

Using the proportion graph, we can see that approximately equal proportion (\~43%) of both biological sexes is sterilized. From the count/column graph, we can see that males are more frequently taken in at the animal shelter. 6352 animals have unknown sex and/or sterilization status.

At this point we could plot `sex` and `sterilisation` in assocation with previously handled variables - `intake_condition` and `animal_type`. I will select only one at this point and then mode on to other variables. I choose `animal_type` to see whether animals, especially the most frequent ones, differ in their sterilization rate and sex. I will focus on proportions, as we have clearly established thus far that dogs and cats are by far the most dominant animal types and visualization of other types besides them is much more informative in proportions.

```{r}
longbeach %>%
  count(animal_type, sex) %>%
  ggplot() + 
  aes(animal_type, n, fill = sex) %>%
  geom_col(position = "fill") + 
  #geom_col() + 
  theme_minimal() + 
  labs(y = "Proportion of animals", x = "Animal type", fill = "Sex", title = "Proportion of different sexes for each type of animal \n(Long Beach Animal Shelter, 2017–2024)") + 
  scale_fill_pomological()
```

We can see that in exotic animals, birds, wildlife and others, very high proportion of those groups have unknown sex. It seems reasonable to assume that this could happen due to accurate genitalia identification is much more difficult to do in these groups, compared to cats, dogs and small mammals. In those groups, male and female sex appear to be approximately balanced. About 10% of cats and small mammals have unknown sex. About 60% of dogs are males and the rest are mostly females and unknown representatives in very small percentage.

Let's look at `sterilised`.

```{r}
longbeach %>%
  count(animal_type, sterilised) %>%
  ggplot() + 
  aes(animal_type, n, fill = sterilised) %>%
  geom_col(position = "fill") + 
  #geom_col() + 
  theme_minimal() + 
  labs(y = "Proportion of animals", x = "Animal type", fill = "Sterilised", title = "Proportion of sterilised individuals for each type of animal \n(Long Beach Animal Shelter, 2017–2024)") +
  scale_fill_pomological()
```

High percentage of individuals with unknown sterilization status in exotic, birds, wildlife and other could be expected. We have seen in previous analysis that a noticeable proportion of those animals are wild animals, thus there is probably no record kept about their sterilization status. Additionally, as mentioned before identification of genitalia and related markers is much more difficult in these species. For cats, dogs and small mammals, sterilization rate is around 40-50%. We can infer some information from this.

If animals have already been taken to a shelter before, they were most likely sterilized, as this is a standard procedure during the capture and stay. While approximately a half of the animals therefore seem to have been captured before, or surrendered/escaped by/from the owner, who put them through the procedure, about the half of these animals have not been sterilized, implying this is their first time in the shelter and are arguably very young. I could explore this further by creating tile plots with `intake_type` and count on axes, and `sterilisation` as fill, faceted for animal types. It would also be possible to perform statistical tests (chi-square) to investigate whether there are significant differences in sterilized vs. unsterilized individuals among different intake or animal types.

```{r}
summary(longbeach)
```

*in case data is overwritten, copy here*

```{r}
#longbeach_after_sex <- longbeach
#longbeach <- longbeach_after_sex
```

### Outcome type

I will proceed to the next variable, `outcome_type`. Let's see what are the levels of the variable and how we could collapse, lump them.

I decide for keeping all the outcomes that are at least 2.5% of the complete sample and lump others. in case of condition "shelter, neuter, return" I am not completely sure what it means as different webpages have different interpretations depending on the context. I will therefore collapse this category with "others".

```{r}
longbeach <- longbeach %>% 
  mutate(outcome_type = fct_lump(outcome_type,  prop = .025)) %>%
  mutate(outcome_type = fct_collapse(outcome_type,
                                     Adoption = "adoption",
                                     Rescue = "rescue",
                                     `Return to owner` = "return to owner",
                                     Euthanasia = "euthanasia",
                                     Died = "died",
                                     Transfer = "transfer",
                                     other_level = "Other"))
                                     

longbeach %>%
  count(outcome_type, sort = TRUE)
```

Let's now create a column `adopted`. It will be "Yes" in case of outcome type "adoption" and "No" for all the other outcome types. **At this point, I remove 187 rows from the dataset, as they hame missing values for `outcome_type`.**

```{r}
longbeach <- longbeach %>%
  filter(!is.na(outcome_type)) %>%
  mutate(adopted = ifelse(outcome_type == "Adoption", 1, 0)) %>%
  mutate(adopted = factor(adopted, levels = c(1, 0), labels = c("Yes", "No")))

longbeach <- longbeach %>%
  relocate(adopted, .after = "outcome_type")
```

Let's also create a binary + other category `outcome_type_posneg` where we will collapse `outcome_type` into positive if animal survived, negative if it did not and other for other level(s).

```{r}

longbeach <- longbeach %>%
  mutate(outcome_type_posneg = fct_collapse(outcome_type,
                                     Positive = c("Adoption", "Rescue", "Return to owner", "Transfer"),
                                     Negative = c("Died", "Euthanasia"),
                                     other_level = "Other"))

longbeach %>%
  mutate(diff = (outcome_type_posneg == "Positive") != was_outcome_alive) %>%
  filter(diff) %>%
  count(outcome_type, was_outcome_alive)
```

There was a discrepancy in number of my "positive outcomes" and original "alive" outcome in variable `was_outcome_alive.` I checked what are outcome types in my active dataset that are alive, but are not "positive" in newly created `outcome_type_posneg` column. All of them are in "other" outcome type. Maybe we can filter out others in future visualization and analysis.

OK! Let's finally proceed to plotting `outcome_type` distribution.

```{r}
outcome_type_posneg_colors <- c(
  "Positive" = "#4c9b82",
  "Negative" = "#e15859",
  "Other" = "#ebcb2a")

longbeach %>% 
  group_by(outcome_type) %>%
  summarise(n = n(),
            outcome_type_posneg = first(outcome_type_posneg)) %>%
  mutate(outcome_type = fct_reorder(outcome_type, n)) %>% 
  ggplot(aes(x = outcome_type, y = n, fill = outcome_type_posneg)) + 
  geom_col() + 
  theme_minimal() + 
  coord_flip() + 
  labs(x = "Outcome type", y = "Number of animals", title = "Number of animals by outcome type (Long Beach Animal Shelter, 2017–2024)", fill = "Outcome type") + 
  #scale_y_continuous(breaks = seq(0,15000,2500)) + 
  #scale_fill_brewer(palette = "Spectral", guide = guide_legend(reverse = TRUE)) +
  #scale_fill_dutchmasters(palette = "milkmaid") + 
  #scale_fill_pomological() + 
  #scale_fill_paletteer_d("PrettyCols::Autumn") +
  scale_fill_manual(values = outcome_type_posneg_colors) +
  theme(plot.title.position = "plot",
        plot.caption = element_text(hjust = -0.3),
        legend.position="none")
# Consider removing the legend
```

This could be a nice lollipop plot!

```{r}
longbeach %>% 
  group_by(outcome_type) %>%
  summarise(n = n(),
            outcome_type_posneg = first(outcome_type_posneg)) %>%
  mutate(
    proportion = n / sum(n),  
    label = scales::percent(proportion, accuracy = 0.1),
    outcome_type = fct_reorder(outcome_type, n) 
  ) %>% 
  ggplot(aes(x = n, y = outcome_type)) + 
  geom_segment(aes(x = 0, xend = n, yend = outcome_type, color = outcome_type_posneg), linewidth = 5) +
  geom_point(aes(color = outcome_type_posneg), size = 13) +
  geom_text(aes(label = label), color = "#313131", size = 3, fontface = "bold") +
  labs(
    y = NULL, x = "Number of animals",
    title = "Proportion of animals by outcome type (Long Beach Animal Shelter, 2017–2024)",
    subtitle = "Outcomes with animal remaining alive and considered positive in green.\nNegative outcomes with animals dying in red.",
    color = "Outcome type",
    caption = "\n2333 individuals in group 'other' were originally considered as living outcome. \nHowever after our recoding they were recoded to neither positive nor negative outcome"
  ) + 
  xlim(0, 7100) + 
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.caption = element_text(hjust = 0),
    legend.position = "none"
  ) + 
  scale_color_manual(values = outcome_type_posneg_colors)
```

### Combining animal type, intake condition and outcome type

Data allows me to plot more specific associations that I am looking to investigate with regression models. First, we will do three things with broader `outcome_type` variable:

-   Visualize animal type vs. outcome type
-   Visualize intake condition vs. outcome type
-   Tile plot - intake condition vs. animal type, values = proportions with positive outcome

Let's get right to it!

#### Animal type vs. outcome type

Let's first determine color scale, that will be used for outcomes in all visualizations equally.

```{r}
outcome_type_colors <- c(
  "Adoption" = "#4c9b82",
  "Return to owner" = "#6cc08b",
  "Euthanasia" = "#e76254",
  "Died" = "#ef8a48",
  "Rescue" = "#90d890",
  "Transfer" = "#d3f2a3",
  "Other" = "#acacab")
```

Animal type vs. outcome type - investigate whether certain animal types are associated with specific outcomes (e.g. cats getting adopted, dogs being transferred etc.)

```{r}
longbeach %>%
  count(animal_type, outcome_type) %>%
  mutate(animal_type = fct_reorder(animal_type, n, .fun = sum)) %>%
  mutate(outcome_type = fct_relevel(outcome_type,
                                    "Euthanasia",
                                    "Died",
                                    "Other",
                                    "Transfer",
                                    "Rescue",
                                    "Return to owner",
                                    "Adoption")) %>%
  ggplot() + 
  aes(animal_type, n, fill = outcome_type) +
  geom_col(position = "fill") + 
  coord_flip() + 
  theme_minimal() + 
  labs(x = "Animal type", y = "Proportion of animals with a certain outcome (%)", fill = "Outcome type", title = "Proportions of different outcomes for different animal types (Long Beach Animal Shelter, 2017–2024") + 
  scale_y_continuous(
  labels = label_percent(), 
  breaks = seq(0, 1, 0.2),
  sec.axis = dup_axis(labels = label_percent(), name = NULL)) +
  theme(
    plot.title.position = "plot",
    plot.caption = element_text(hjust = 0)) +
  scale_fill_manual(values = outcome_type_colors, guide = guide_legend(reverse = TRUE))
```

There are a few things we can conclude from this visualization. Small mammals have the highest proportion of adoptions, followed closely by dogs and cats. While a insignificant proportion of cats is returned to the owners, situation is different for dogs. About 30% of dogs are returned to their owner. It is possible that a significant percentage of stray dogs have escaped their home and were roaming free for a while and then returned to their owner. On the other hand it is possible that much lower proportion of stray cats have escaped their home recently, but are actually street-based animals. Overall, proportion of animals euthanized is about 2.5x higher for cats than for dogs. Highest proportion of euthanasia in a specific animal type is in wildlife (\~60%) and animals classified as "other"(\~50%). We have speculated in the previous chapters that wildlife animals perhaps end up in the shelter more frequently because of their high rate of moderate and severe injuries (50-55% combined; reported in section "Combining animal type and intake condition"). It is possible that if big proportion of these animals cannot get better, be healed, they highly likely end up euthanized instead of returned to natural habitat.

While positive outcomes (bar segments in green) are about 90% for dogs, 80% for small mammals, 80% for exotic animals, 65% for cats and slightly above 60% for birds, wildlife animals and others are more likely to result in a negative outcome (55% for others, 65% for wildlife).

#### Intake condition vs. outcome type

I will now repeat the same procedure as in the previous section to invastigate possible associations between intake condition and outcome type. With this, we will be able to get some information whether the condition in which the animal is taken into the shelter can give the caretakers any approximation about the most likely outcome for that animal.

```{r}
longbeach %>%
  count(intake_condition, outcome_type) %>%
  mutate(intake_condition = fct_relevel(intake_condition,
                                        "Normal",
                                        "Under age/weight",
                                        "Mild physical",
                                        "Moderate physical",
                                        "Severe physical",
                                        "Other")) %>%
  mutate(outcome_type = fct_relevel(outcome_type,
                                    "Euthanasia",
                                    "Died",
                                    "Other",
                                    "Transfer",
                                    "Rescue",
                                    "Return to owner",
                                    "Adoption")) %>%
  ggplot() + 
  aes(intake_condition, n, fill = outcome_type) +
  geom_col(position = "fill") + 
  coord_flip() + 
  theme_minimal() + 
  labs(x = "Intake", y = "Proportion of animals (%)", fill = "Outcome type", title = "Proportions of specific outcome for different intake conditions (Long Beach Animal Shelter, 2017–2024)") + 
  scale_y_continuous(
  labels = label_percent(), 
  breaks = seq(0, 1, 0.2),
  sec.axis = dup_axis(labels = label_percent(), name = NULL)) +
  theme(
    plot.title.position = "plot",
    plot.caption = element_text(hjust = 0)) +
  scale_fill_manual(values = outcome_type_colors, guide = guide_legend(reverse = TRUE))
```

From this plot we can see that intake condition is associated with the outcome. Almost 90% of animals taken in in normal condition are likely to face a positive outcome, meaning they will stay alive and either get adopted, returned to their owner or be transferred to another rescue or shelter facility. With increase in severity of physical condition, the chances of positive outcome decrease, with about 70% positive outcomes for under-age and/or under-weight animals, and those with mild physical condition. About 55% of animals in moderate physical condition will face a positive outcome and this proportion drops to about 25% for severely injured of ill animals.

**NOTE:** Before proceeding, I will remove category "Other" from `animal_type` and `intake_condition`, to avoid repeating it within every pipeline. 

```{r}
# In case copy of longbeach is needed in the future - before filtering animal_type and intake_condition for "Others"

#longbeach_before_fitlering_Others <- longbeach
#longbeach <- longbeach_before_fitlering_Others
```

Filter:
```{r}
longbeach <- longbeach %>% 
  filter(intake_condition != "Other" & animal_type != "Other")
```


#### Animal type vs. intake condition vs. outcome type

Let's now connect all three variables in a tile plot. I will plot animal type against intake condition and use gradient of the tiles to show what is the percentage of positive vs negative outcome. We will not use specific `outcome_type` categories, but return to the variable `outcome_type_posneg`. As a reminder, we collapsed `outcome_type` categories into positive outcome if animal survived, negative if it did not and other for other level(s). Positive outcomes correspond with green segments and negative outcomes with red in previous graphs in this section.

Note: proportion was calculated as positive / positive + negative + other. Therefore it is crucial to notice that proportional graphs tell us about proportion of positive outcomes compared to negative and other outcomes, not just negative. Other outcomes are significantly lower in count compared to negative however

```{r}
longbeach %>% 
  # count outcome +/- based on intake condition for every animal
  count(animal_type, intake_condition, outcome_type_posneg) %>%                       
  group_by(animal_type, intake_condition) %>%
  #calculate total outcomes for each income condition for each animal type and proportion (e.g., positive outcomes for normal cats/ all outcomes for normal cats)
  mutate(outcome_type_total_posnegother = sum(n),
         positive_outcome_type_prop = n/outcome_type_total_posnegother) %>%
  ungroup() %>%
  # add total number of ouccomes/cases for every animal type
  group_by(animal_type) %>%
  mutate(total_animal_count = sum(n)) %>%
  ungroup() %>%
  # add total number of outcomes/cases for every intake condition
  group_by(intake_condition) %>%
  mutate(intake_condition_total = sum(n)) %>%
  ungroup() %>%
  # Add new column with animal types and with their total count in brackets and relevel/reorder them in the same step
  mutate(animal_type_label = fct_reorder(paste0(animal_type, " (", total_animal_count, ")"),
                                         total_animal_count)) %>%
  mutate(intake_condition = fct_relevel(intake_condition,
                                        "Normal",
                                        "Under age/weight",
                                        "Mild physical",
                                        "Moderate physical",
                                        "Severe physical",
                                        "Other")) %>%
  filter(outcome_type_posneg == "Positive") %>%
  ungroup() %>%
  ggplot() + 
  aes(x = animal_type_label, y = intake_condition, fill = positive_outcome_type_prop*100) +
  geom_tile(color = "white") + 
  geom_text(aes(label = paste0(n, "\n(", outcome_type_total_posnegother, ")")), color = "#313131", size = 3) + 
  coord_flip() +
  #coord_fixed() + 
  scale_fill_gradientn(
    colors = c("#e76254", "#ef8a48", "#fbd587", "white","#d3f2a3", "#90d890", "#4c9b82"),
    labels = scales::label_percent(scale = 1)) +
  labs(x = NULL, y = "Intake condition of the animal", title = "Positive outcome proportion for each animal type and intake condition (Long Beach Animal Shelter, 2017–2024)", subtitle = "Numbers on tiles represent total number of positive outcomes for that intake condition in each animal type.\nNumber in paranthesis is total number of individuals from that animal type with specific intake condition.", fill = "Positive outcome rate \n") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title.position = "plot",
    plot.caption = element_text(hjust = 0))
  
```

There are several important pieces of information we can gather from this plot. First thing to highlight is that across different animal types we can see decrease in positive outcome proportion between mild and severe physical conditions, as gradient goes from greenish to reddish colors. This is especially noticeable in cats, dogs and small mammals. However, it seems that physical conditions have a more pronounced effect in cats compared to dogs. Generally a lower proportion of positive outcomes is observed across different physical conditions for cats compared to dogs. In general, dogs appear to have the highest positive outcome rate, as more than 50% positive outcomes are expected across all conditions except severe physical. On the other hand, wildlife animals face a positive outcome less than 50% of time across all intake conditions, as indicated by dominantly orange-red colored tiles.

```{r}
longbeach %>% 
  count(animal_type)
```

Let's now repeat this whole section, using adoption instead of broader outcome types.

### Combining animal type, intake condition and adoption

In this section I will focus on replicating the last plot (tile plot) from the last section. It would not be that interesting to plot `animal_type vs adoption` and `intake_condition vs adoption`, because filling the columns by outcome type in the previous section gave us explicit information about adoptions, as one of the column fills referred to this category specifically (dark green on previous plots).

```{r}
longbeach %>% 
  # count outcome +/- based on intake condition for every animal
  count(animal_type, intake_condition, adopted, .drop = FALSE) %>%                      
  group_by(animal_type, intake_condition) %>%
  #calculate total outcomes for each income condition for each animal type and proportion (e.g., positive outcomes for normal cats/ all outcomes for normal cats)
  mutate(outcome_type_total = sum(n),
         outcome_rate = n/outcome_type_total,
         outcome_rate = ifelse(is.na(outcome_rate), 0, outcome_rate)) %>%
  ungroup() %>%
  # add total number of outcomes/cases for every animal type
  group_by(animal_type) %>%
  mutate(total_animal_count = sum(n)) %>%
  ungroup() %>%
  # add total number of outcomes/cases for every intake condition
  group_by(intake_condition) %>%
  mutate(intake_condition_total = sum(n)) %>%
  ungroup() %>%
  # Add new column with animal types and with their total count in brackets and relevel/reorder them in the same step
  mutate(animal_type_label = fct_reorder(paste0(animal_type, " (", total_animal_count, ")"),
                                         total_animal_count)) %>%
  mutate(intake_condition = fct_relevel(intake_condition,
                                        "Normal",
                                        "Under age/weight",
                                        "Mild physical",
                                        "Moderate physical",
                                        "Severe physical",
                                        "Other")) %>%
  filter(adopted == "Yes" & intake_condition != "Other" & animal_type != "Other") %>%
  ungroup() %>%
  ggplot() + 
  aes(x = animal_type_label, y = intake_condition, fill = outcome_rate*100) +
  geom_tile(color = "white") + 
  geom_text(aes(label = paste0(n, "\n(", outcome_type_total, ")")), color = "#313131", size = 3) + 
  scale_fill_gradientn(
    colors = c("#e76254", "#ef8a48", "white","#d3f2a3", "#90d890", "#4c9b82"),
    labels = scales::label_percent(scale = 1)) +
  #coord_fixed() + 
  labs(x = NULL, y = "Intake condition of the animal", title = "Adoption rate for each animal type and intake condition (Long Beach Animal Shelter, 2017–2024)", subtitle = "Numbers on tiles represent total number of adoptions for that intake condition in each animal type.\nNumbers in paranthesis are total number of individuals from that animal type with specific intake condition.", fill = "Adoption rate") + 
  theme_minimal() +
  coord_flip() +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title.position = "plot",
    plot.caption = element_text(hjust = 0))
```

Similarly to the previous plot, this one reveals several interesting details about how intake condition is associated with adoption rate in different animal types. Highest rates of adoptions are in normal and mildly physically compromised cats and small mammals. This is followed by normal and mildly physically compromised dogs. We can see a severe decrease in adoption rate from mild to severe physical condition in dogs and cats. In all groups with the exception of cats, under-age and under-weight animals seem to have little to no chances of getting adopted. This suggest that potential future owners possibly believe that raising such an animal will be very difficult, or possibly shelters do not offer these animals for adoption as often. Though in the section "Intake condition vs. outcome type" we could see that about 15% of under-age and/or under-weight animals get adopted. It is possible that majority of those animals are cats, as we can see in this plot tat about 20% of under-age and/or under-weight cats get adopted. As we have seen in the section "Animal type vs. outcome type", no wildilfe animals get adopted. Birds have a similarly low rate with a few exceptions in compromised groups and 26 normal individuals getting adopted. Adoption rates are small for exotic animals as well, as only 11 normal and 1 per each physically compromised category were adopted.

Overall, there seems to be more of a binary difference in animal type trends, as cats, dogs and small mammals have adoption rates close to 25-30% percent. On the other hand, other groups share similarly low adoption rate at maximum 5%. Intake conditions exhibit more stepwise (linearly) decreasing rates from normal to most severely physically compromised group. Importantly however, mildly injured animals, especially cats and dogs show higher adoption rates compared to under-age and -weight group. This could be due to fewer offerings of these animals by shelter. On the other hand potential adopters might believe that it is easier to help get mildly compromised animal better, than to take care of under-age or -weight individual in the long run.


///////

Ok, let's try to improve the plot and then separately create the table, so that the plot is not oversaturated with information. 

Plot
```{r}
# Prepare the order of the levels according tow hat we want in the final plot - animal names sorted by total number and conditions by severity

condition_pcts_labels <- longbeach %>% 
  count(intake_condition, adopted, .drop = FALSE) %>%
  group_by(intake_condition) %>%
  mutate(total_per_condition = sum(n)) %>%
  filter(adopted == "Yes") %>%
  mutate(condition_adopted_pct = n/total_per_condition,
         intake_condition_label = paste0(intake_condition, "\n(", round(condition_adopted_pct*100,1), "%)")) %>%
  filter(intake_condition != "Other") %>%
  pull(intake_condition_label)

condition_pcts_labels <- condition_pcts_labels[c(4,5,1,2,3)]

animal_type_pcts_labels <- longbeach %>% 
  count(animal_type, adopted, .drop = FALSE) %>%
  group_by(animal_type) %>%
  mutate(total_per_type = sum(n)) %>%
  filter(adopted == "Yes") %>%
  mutate(type_adopted_pct = n/total_per_type,
         animal_type_label = paste0(animal_type, "\n(", round(type_adopted_pct*100,1), "%)")) %>%
  filter(animal_type != "Other") %>%
  pull(animal_type_label)

animal_type_pcts_labels <- rev(c("Cats\n(25.2%)","Dogs\n(26.7%)" ,"Small mammals\n(30%)","Birds\n(1.5%)","Exotic\n(4.1%)","Wildlife\n(0%)"))

```


Plot:
```{r}
# calculate adoption proportions for every animal type
animal_type_adoption_pcts <- longbeach %>% 
  count(animal_type, adopted, .drop = FALSE) %>%
  group_by(animal_type) %>%
  mutate(total_per_type = sum(n)) %>%
  filter(adopted == "Yes") %>%
  mutate(type_adopted_pct = n/total_per_type,
         animal_type_label = paste0(animal_type, "\n(", round(type_adopted_pct*100,1), "%)")) %>%
  filter(animal_type != "Other") %>%
  select(animal_type, type_adopted_pct, animal_type_label)
  
# calculate adoption proportions for every intake condition
intake_cond_adoption_pcts <- longbeach %>% 
  filter(intake_condition != "Other" & animal_type != "Other") %>%
  count(intake_condition, adopted, .drop = FALSE) %>%
  group_by(intake_condition) %>%
  mutate(total_per_condition = sum(n)) %>%
  filter(adopted == "Yes") %>%
  mutate(condition_adopted_pct = n/total_per_condition,
         intake_condition_label = paste0(intake_condition, "\n(", round(condition_adopted_pct*100,1), "%)")) %>%
  select(intake_condition, condition_adopted_pct, intake_condition_label)
   

# Calculate for each type*condition pair and merge with previous datasets

longbeach %>% 
  # count outcome +/- based on intake condition for every animal
  count(animal_type, intake_condition, adopted, .drop = FALSE) %>%                
  group_by(animal_type, intake_condition) %>%
  #calculate total outcomes for each income condition for each animal type and proportion (e.g., positive outcomes for normal cats/ all outcomes for normal cats)
  mutate(anicond_total_outcomes = sum(n),
         anicond_outcome_rate = n/anicond_total_outcomes,
         anicond_outcome_rate = ifelse(is.na(anicond_outcome_rate), 0, anicond_outcome_rate)) %>%
  ungroup() %>%
  filter(adopted == "Yes" & intake_condition != "Other" & animal_type != "Other") %>%
  left_join(animal_type_adoption_pcts, by = "animal_type") %>%
  left_join(intake_cond_adoption_pcts, by = "intake_condition") %>%
  mutate(animal_type_label = factor(animal_type_label, levels = animal_type_pcts_labels),
         intake_condition_label = factor(intake_condition_label, levels = condition_pcts_labels)) %>%
    ggplot() + 
  aes(x = animal_type_label, y = intake_condition_label, fill = anicond_outcome_rate*100) +
  geom_tile(color = "white") + 
  geom_text(aes(label = paste0(n, "\n(", anicond_total_outcomes, ")")), color = "#313131", size = 3) + 
  scale_fill_gradientn(
    colors = c("#e76254", "#ef8a48", "white","#d3f2a3", "#90d890", "#4c9b82"),
    labels = scales::label_percent(scale = 1)) +
  labs(x = NULL, y = NULL, title = "Adoption rate for each animal type and intake condition (Long Beach Animal Shelter, 2017–2024)", subtitle = "Numbers on tiles represent total number of adoptions for that intake condition in each animal type.\nNumbers in paranthesis are total number of individuals from that animal type with specific intake condition.", fill = "Adoption rate") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title.position = "plot",
    plot.caption = element_text(hjust = 0)) +
  coord_flip() +
  theme(aspect.ratio = 0.66)
```


Table:
```{r}
# Maybe not needed since previous graphs includes both image info and text......will think about it. 
```


### Age

Let's now focus on the only continuous variable that I am planning to use in the analysis. Let's first check the type of variable and if there are any false data entry points.
```{r}
summary(longbeach)
# As dob and intake_date are now "double", we have to change them to date variables
longbeach <- longbeach %>%
  mutate(dob = as.Date(dob),
         intake_date = as.Date(intake_date))

# Now let's see what is the distribution of these dates and if there are any false datapoints 
summary(longbeach)

# Well there is a date set in 2031 in dob variable, which cannot be as we are only in 2025. Let's remove this value from the dataset.
longbeach <- longbeach %>%
  filter(dob < as.Date("2024-12-31"))

longbeach %>%
  ggplot(aes(dob)) + 
  geom_histogram() + 
  labs(x = "Date of birth", y = "Number of animals", title = "Distribution of animals in the shelter by date of birth \n(Long Beach Animal Shelter, 2017–2024)") + 
  theme_minimal()
```

We can now use `dob` and `intake_date` to calculate the age on an animal at intake, which I will use later on in the regression models.
```{r}
longbeach %>%
  mutate(Age = ifelse(is.na(dob), NA, intake_date-dob)) %>%
  filter(Age > 0) %>%
  ggplot() + 
  aes((Age), fill = adopted) + 
  geom_histogram(bins = 20, position = "dodge") + 
  scale_x_log10() + 
  theme_minimal() + 
  labs(x = "Age in days (log transformed)", y = "Number of animals", title = "Adoption outcomes per age on an animal \n(Long Beach Animal Shelter, 2017–2024)", fill = "Adopted")
  #facet_wrap(~adopted)
```
```{r}
longbeach %>%
  mutate(Age = as.numeric(intake_date - dob)) %>%
  filter(!is.na(adopted), Age > 0) %>%
  mutate(adopted_num = ifelse(adopted == "Yes", 1, 0)) %>%
  ggplot(aes(x = Age, y = adopted_num)) +
  geom_smooth(method = "loess", formula = y ~ x, se = TRUE) +
  #scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  labs(x = "Age in days (log transformed)", y = "Adoption Rate", title = "Adoption Rate by Age") + 
  theme_minimal() + 
  scale_x_log10() 
```


```{r}
longbeach %>%
  filter(animal_type == "Cats" | animal_type == "Dogs") %>%
  mutate(Age = as.numeric(intake_date - dob)) %>%
  filter(!is.na(adopted), Age > 0) %>%
  mutate(adopted_num = ifelse(adopted == "Yes", 1, 0)) %>%
  ggplot(aes(x = Age, y = adopted_num)) +
  geom_smooth(method = "loess", formula = y ~ x, se = TRUE) +
  #scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  labs(x = "Age in days (log transformed)", y = "Adoption Rate", title = "Adoption Rate by Age") + 
  theme_minimal() + 
  scale_x_log10() + 
  facet_wrap(~animal_type)
```

Huh, alright. At this point I have to decide how to continue with this variable. I see two options - one is log() + quadratic model, as visualisation suggest rughly a quadratic relationship between age and adoption rate. I do not have that much experience with the combination of these, and while I know how to construct a model, I am a bit reserved about the visualizations (still feasible) and model interpretation. I am thinking to use grouping / binning for `Age` and use corresponding categories for different animal types. Binning a continuous variable into groups leads to a loss of statistical power. This is because binning reduces the amount of information available for analysis by grouping continuous values into discrete categories, making it harder to detect subtle but real relationships between variables. Though I still believe at this point, for a solo UNI assignment, where I do not have collaborators to work together on result/model interpretation etc. I think I can do my work better with grouped variable. 


# Logistic regression models

In this analysis I aim to investigate whether contextual factors—such as the condition in which the animal was taken in and whether it was sterilised—can significantly improve the prediction of adoption outcomes, compared to a model that includes only animal-focused characteristics.

**Research Question:**
Does the inclusion of contextual variables improve model fit in predicting whether an animal gets adopted?

**Hypothesis:**
Based on preliminary exploratory analysis, I hypothesise that contextual factors will significantly improve the model’s predictive performance over a baseline model that includes only animal-level variables.

**Data Description:**

The outcome variable is Adoption (1 = adopted, 0 = not adopted).

I restricted the dataset to cats and dogs, excluding other animal types such as birds, exotic, and wildilfe animals, and small mammals due to insufficient sample sizes and for practical reasons that dogs and cats are usually most prevalent animal types in shelters and are most commonly adopted animal types. This narrows the investigation and focuses on specific groups and their adoption rates.

**SIMPLE MODEL **

In the simple model, I will focus on animal-related, native, given factors, which do not depend on contextual factors or interventions. They include:

- Animal Type (cat or dog)
- Sex (male or female)
- Age, categorized into five bins reflecting developmental stages:

  -Junior (0–1 year),
  -Young Adult (1–6 years for cats, up to 2 years for dogs),
  -Mature Adult (7–10 years for cats, 3–6 years for dogs),
  -Senior (10+ for cats, 7–11 years for dogs),
  -Geriatric (older then previous levels; in both categories).

This categorization reflects species-specific aging and was used to simplify interpretation and ensure comparability across species. Categorization was based on scientific guidelines for age categorization (Harvey, 2021; Quimby et al., 2021).

**COMPLEX MODEL**
In the complex model I will add so called contextual predictors, which are situational or depend on interventions. They will include:

- Intake condition, reduced to five categories (normal, under age/weight, mild physical, moderate physical, severe physical). *Less frequent or poorly defined were excluded due to small sample sizes and lack of sufficient context for interpretation.*
- Sterilized status (1 = yes, 0 = no)


These variables were selected based on theoretical relevance, availability, and data quality. The regression analysis compares a baseline model including only animal-level variables to a complex model that incorporates contextual predictors.

## Final dataset preparation
```{r}
lb_analysis <- longbeach %>%
  # Use only cats and dogs
  filter(animal_type == "Cats" | animal_type == "Dogs") %>%
  # Filter out other intake conditions due to reasons listed above
  filter(intake_condition != "Other") %>%
  # Filter out Unknown sterilisation status
  filter(sterilised != "Unknown") %>%
  mutate(age_years = as.numeric(intake_date - dob)/365) %>%
  # Filter false data, as there are false datapoints (age < 0)  (113 datapoints removed)
  filter(age_years > -0.0001) %>%   
  # Grouping into age bins
  mutate(age_bins = case_when(
    # Junior (0–1 year) for both cats and dogs
    age_years < 1 ~ "Junior",
    
    # Cats
    animal_type == "Cats" & age_years >= 1 & age_years <= 6 ~ "Young adult",
    animal_type == "Cats" & age_years > 6 & age_years <= 10 ~ "Mature adult",
    animal_type == "Cats" & age_years > 10 & age_years <= 13 ~ "Senior",
    animal_type == "Cats" & age_years > 13 ~ "Geriatric",
    
    # Dogs
    animal_type == "Dogs" & age_years >= 1 & age_years < 2 ~ "Young adult",
    animal_type == "Dogs" & age_years >= 2 & age_years <= 6 ~ "Mature adult",
    animal_type == "Dogs" & age_years > 6 & age_years <= 11 ~ "Senior",
    animal_type == "Dogs" & age_years > 11 ~ "Geriatric",
    
    TRUE ~ NA_character_
  )) %>%
  mutate(age_bins = fct_relevel(age_bins, "Junior", "Young adult", "Mature adult", "Senior", "Geriatric"),
         intake_condition = fct_relevel(intake_condition, "Normal", "Mild physical", "Moderate physical", "Severe physical", "Under age/weight"),
         adopted = ifelse(adopted == "Yes", 1, 0),
         sterilised = factor(sterilised),
         sex = factor(sex)) %>%
  select(animal_id, animal_type, sex, age_bins, intake_condition, sterilised, adopted)

```



## Simple model (animal related, native factors)

### Construct the model
```{r}
lb_simple_mdl <- glm(adopted ~ animal_type + sex + age_bins, 
                  data = lb_analysis, family = binomial)
```


Check model summary:
```{r}
summary(lb_simple_mdl)
```

And model performance:
```{r}
model_performance(lb_simple_mdl)
```


### Visualize the model
```{r}
plot_model(lb_simple_mdl, type = "pred", terms = c("age_bins", "animal_type")) + theme_minimal() + labs(x = "Age bins", y = "Adopted", title = "Predicted probabilities of adoption for dogs and cats, separated by age bins", color = "Animal type")
```

### Checking assumptions
```{r}
check_model(lb_simple_mdl)
```
Check residuals:
```{r}
check_residuals(lb_simple_mdl)
```

Multicolinearity:
```{r}
check_collinearity(lb_simple_mdl)
```

Influential points:
```{r}
# Leverage
n_obs = nrow(lb_analysis)
params = length(coef(lb_simple_mdl)) - 1

influential_lev <- hatvalues(lb_simple_mdl) >  2*(params+1)/n_obs
sum(influential_lev)

lb_simple_mdl %>%
  augment() %>%
  mutate(high_leverage = ifelse(.hat > 2*(params+1)/n_obs, 1, 0)) %>%
  count(high_leverage)

# OK so there are 781 influential points if we only consider leverage. Let's now check Cook's distance.

# Cook's distance
influential_cd <- cooks.distance(lb_simple_mdl) > 4 / n_obs
sum(influential_cd) 
sum(cooks.distance(lb_simple_mdl) > 1)
# And 523 influential points according to 4/n rule of thumb and 0 according to D > 1. 

#First I want to check whether leverage and Cook's distance influential points overlap
check_influence_overlap = data.frame(lev = influential_lev, cd = influential_cd )
check_influence_overlap %>%
  summarise(sum(lev == TRUE & cd == TRUE))
# There are 118 points that are considered influential both to leverage and Cook's distance threshold

# I want to check the data that is influential according to Cook's distance
lb_analysis %>%
  filter(influential_cd) %>%
  count(age_bins)

# All the influential points according to Cook's distance come from older groups. Let's now check what is overall number of adoptions per age bin. 

lb_analysis %>%
  #filter(influential_cd) %>%
  count(age_bins, adopted)

# Hm, ok. We see that basically all the influential points from "geriatric" and "senior" are datapoints that are also "adopted".More on this topic in the text below. 
```

Cook’s distance analysis revealed influential observations (n = 523) concentrated among adopted senior (n=336/1,802 total seniors) and geriatric (n=67/586 total geriatrics) animals. Notably, these points accounted for all adoptions in their age groups—when hypothetically excluded, the model would lose its ability to estimate adoption effects for senior/geriatric animals entirely. To address this,I will try one solution: merge senior and geriatric categories into an "Older" group and re-analyse the model to see what happens. 


## Updated simple model (merging "geriatric" and "senior" age bins)

Merge levels:
```{r}
lb_analysis_merged_age <- lb_analysis %>%
  mutate(age_bins = fct_collapse(age_bins, "Senior" = c("Senior", "Geriatric")))
```


Construct new model:
```{r}
lb_simple_merged_mdl <- glm(adopted ~ animal_type + sex + age_bins, 
                  data = lb_analysis_merged_age, family = binomial)
```

Check model summary:
```{r}
summary(lb_simple_merged_mdl)
```

And model performance:
```{r}
model_performance(lb_simple_merged_mdl)
```


### Visualize the model
```{r}
plot_model(lb_simple_merged_mdl, type = "pred", terms = c("age_bins", "animal_type")) + theme_minimal() + labs(x = "Age bins", y = "Adopted", title = "Predicted probabilities of adoption for dogs and cats, separated by age bins", color = "Animal type")
```


### Checking assumptions

Multicolinearity:
```{r}
check_collinearity(lb_simple_merged_mdl)
```

Influential points:
```{r}
# Leverage
n_obs_2 = nrow(lb_analysis)
params_2 = length(coef(lb_simple_merged_mdl)) - 1

influential_lev_2 <- hatvalues(lb_simple_merged_mdl) >  2*(params_2+1)/n_obs_2
sum(influential_lev_2)

lb_simple_merged_mdl %>%
  augment() %>%
  mutate(high_leverage = ifelse(.hat > 2*(params_2+1)/n_obs_2, 1, 0)) %>%
  count(high_leverage)

# OK so after merging age bins, there are no more influential points when leverage is considered.

# Cook's distance
influential_cd_2 <- cooks.distance(lb_simple_merged_mdl) > 4 / n_obs_2
sum(influential_cd_2) 
sum(cooks.distance(lb_simple_merged_mdl) > 1)
# We still have 523 influential points according to 4/n rule of thumb and 0 according to D > 1. However maxmimum Cook's distance decreased from about 0.002 to about 0.0005, so there was definitely some improvement after age bin merging. 

```

Check residuals - here is where I ran into a bit of confusion that will probably take me some time to figure out, probably more time is needed than the deadline for the report. I was checking residuals in 2 ways, one with `check_residuals` and another with `binned residuals` (see both below). 

Initial checks with check_residuals() indicated no major violations of model assumptions (quantile residuals followed expected uniform distribution, however this is NOT a requirement in any case, at least to my current knowledge). However, the binned residuals plot (binned_residuals()) showed poor fit, with only 31% of residuals within error bounds—suggesting localized prediction inaccuracies for certain subgroups. While tools like DHARMa (simulated residuals) were explored, the results (a gray-shaded plot with a straight red line) proved difficult to interpret given time constraints and my limited familiarity with advanced diagnostics. At this point I am not sure how to deal with this problem, so I think I will continue and be careful about interpretation of model results later on. 

```{r}
check_residuals(lb_simple_merged_mdl)
binned_residuals(lb_simple_merged_mdl)
simulated_residuals <- simulateResiduals(lb_simple_merged_mdl)
plot(simulated_residuals)
```








## Complex model

```{r}
lb_complex_mdl <- glm(adopted ~ animal_type + sex + age_bins + sterilised + intake_condition, 
                  data = lb_analysis, family = binomial)


```{r}
summary(lb_complex_mdl)
```

```{r}
plot_model(lb_complex_mdl, type = "pred", terms = c("animal_type", "intake_condition", "age_bins", "sterilised [Yes]"))
```
```{r}
plot_model(lb_complex_mdl, type = "pred", terms = c("animal_type", "intake_condition", "age_bins", "sterilised"))
```



```{r}
r2(lb_complex_mdl)
check_model(lb_complex_mdl)
binned_residuals(lb_complex_mdl)

library(DHARMa)
simulated_residuals <- simulateResiduals(lb_complex_mdl)
plot(simulated_residuals)
```

```{r}
lb_simple_mdl %>%
  augment() %>%
  arrange(desc(.cooksd)) %>%
  mutate(cd_t = ifelse(.cooksd > 4/n(), "Influential", "Ok")) %>%
  count(cd_t)

plot(lb_simple_mdl, which = 4)
influence.measures(lb_simple_mdl)
```

Let's see what happens if we remove outliers.

```{r}
# Check influential points
influential <- cooks.distance(lb_simple_mdl) > 4 / nobs(lb_simple_mdl)
sum(influential)  # ~500

lb_analysis %>%
  filter(influential) %>%
  count(age_bins)

lb_analysis %>%
  count(age_bins, intake_condition)

# Compare models
simple_model_clean <- update(lb_simple_mdl, subset = !influential)
summary(lb_simple_mdl)
summary(simple_model_clean)


plot_model(simple_model_clean, type = "pred")
coef_diff  # Examine changes
```


### Model comparisons

```{r}
anova(lb_simple_mdl, lb_complex_mdl)
compare_performance(lb_simple_mdl, lb_complex_mdl, rank = TRUE, verbose = FALSE)
```

```{r}
test_performance(lb_simple_mdl, lb_complex_mdl)
```

