---
title: "Long Beach Animal Shelter - EDA"
author: "Miha Likar"
format: html
editor: visual
---

# Introduction

In this report, I'll present exploratory data analysis of Long Beach Animal Shelter dataset. Additionally, throughtout the EDA, I will formulate specific research question(s) which will be addressed in the regression models section. 

The workflow and structure of this report is as follows:

- Brief introduction about the dataset and available variables
- Explore distributions of various single variables
- Explore associations between 2 and/or more variables
- Impute certain variables based on the available data (age, reproductive status)
- Clearly formulate RQ and Hs
- Regression analysis (model comparison)






When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
